---
title: "Hair-Loss-EDA-Statistical-Analysis"
author: "Xavier Tarragó Claramunt"
date: "2025-28-10"
output: html_document
---

## ***Data Import, Export and Management***

```{r}

# Load required libraries:
library(readr)
library(knitr)

# Read the file and suppress column type messages
HairFallH <- read_csv("Predict Hair Fall.csv") 

# Replace all cells containing "No Data" with NA
HairFallH[HairFallH == "No Data"] <- NA

# Display the first 6 rows as a table
kable(head(HairFallH)) 
```


```{r}
# Create a data frame summarising the dataset characteristics
resum <- data.frame(
  Data_type  = sapply(HairFallH, class),                          # Data type of each variable
  Size       = sapply(HairFallH, function(x) sum(!is.na(x))),     # Number of non-NA observations
  Null_values = sapply(HairFallH, function(x) sum(is.na(x))))     # Number of NA values

# Display as a table
kable(resum)
```


### Proportion of people by stress level, grouped by type of nutritional deficiency

```{r}
# Remove NA values before any operation
HairFallH_clean <- na.omit(HairFallH)

# Verify that all NAs have been removed
any(is.na(HairFallH_clean))

# Create the contingency table
taula <- table(`Nutritional Deficience` = HairFallH_clean[["Nutritional Deficiencies"]],
               `Stress` = HairFallH_clean$Stress)

# Compute row proportions (percentage of each stress level within each nutritional deficiency)
prop_tabla <- prop.table(taula, margin = 1)  

# Round proportions to 2 decimal places
round(prop_tabla, 2)
```

Deficiencies such as Protein and Vitamin E show a higher proportion of high stress and a lower proportion of low stress, suggesting a possible association with elevated stress levels. In contrast, deficiencies such as Biotin and Omega-3 are more common among people with low stress, indicating a link with lower stress. Some deficiencies, such as Selenium and Magnesium, show a higher proportion of moderate stress, reflecting a more balanced stress distribution. The table allows us to observe how different nutritional deficiencies relate to stress levels, although it does not establish causality.

### Relationship between stress and hair loss

```{r}

# Recode 0/1 values with descriptive labels
HairFallH_clean$Hair_Loss_Label <- ifelse(HairFallH_clean$`Hair Loss` == 0, "No Hair Loss", "Hair Loss")

library(ggplot2)

# Bar chart comparing stress levels by hair loss status
ggplot(HairFallH_clean, aes(x = Hair_Loss_Label, fill = Stress)) +
  geom_bar(position = "dodge") +
  labs(title = "Proportion of stress levels according to Hair Loss",
       x = "Hair Loss", y = "Frequency") +
  theme_minimal()

```

Based on the results, no clear difference is observed between age and stress. The trend remains similar in both groups, with moderate stress being the most common in both cases.

### Relationship between hair loss and specific nutritional deficiencies, counting the number of people per deficiency type

```{r}

# Create the contingency table
taula <- table(HairFallH_clean$`Nutritional Deficiencies`, HairFallH_clean$Hair_Loss_Label)

# Display as a table
kable(taula)

```

### Number of people with hair loss, grouped by stress level, who smoke and are below the mean age

```{r}
# Calculate the mean age
mean_age <- mean(HairFallH_clean$Age)
mean_age

# Filter people meeting all specified conditions
Filtrat <- HairFallH_clean[
  HairFallH_clean$Hair_Loss_Label == "Hair Loss" &
  HairFallH_clean$Smoking == "Yes" &
  HairFallH_clean$Age < mean_age, 
]

# Display the table grouped by stress level
taula_df <- as.data.frame(table(Filtrat$Stress))
colnames(taula_df) <- c("Stress level", "Number of people")

kable(taula_df)

```


### Nutritional deficiency among the youngest people with hair loss, who smoke, have high stress and are under medical treatment

```{r}

# Find the minimum age among people meeting all specified conditions
subset_filtrat <- HairFallH_clean[
  HairFallH_clean$Hair_Loss_Label == "Hair Loss" &
  HairFallH_clean$Smoking == "Yes" &
  HairFallH_clean$Stress == "High" &
  HairFallH_clean$`Medications & Treatments` != "",   # Ensure the field is not empty
]
  
# Find the minimum age
min_age <- min(subset_filtrat$Age)

# Select people with this minimum age
youngest <- subset_filtrat[subset_filtrat$Age == min_age, ]

# Display their nutritional deficiencies
kable(youngest[, c("Age", "Hair_Loss_Label", "Smoking", "Stress", "Medications & Treatments", "Nutritional Deficiencies")])

```

In conclusion, the minimum age among people meeting these conditions is 24 years. Three individuals present hair loss, smoke, have high stress and are under medical treatment; two have Vitamin A deficiency and one has Vitamin D deficiency.

### Most frequent medical conditions among people with hair loss, and how this frequency varies by the presence or absence of environmental factors

```{r, message=FALSE, warning=FALSE}

# Load required libraries:
# dplyr for data manipulation, ggplot2 for visualisation
library(dplyr)
library(ggplot2)

# Filter only people with hair loss
HairFall_FT <- HairFallH_clean %>%
  filter(Hair_Loss_Label == "Hair Loss") %>%
  filter(`Medical Conditions` != "", `Environmental Factors` != "")

# Count people with hair loss for each medical condition by environmental factor
taula <- HairFall_FT %>%
  count(`Medical Conditions`, `Environmental Factors`)

# Bar chart
ggplot(taula, aes(x = reorder(`Medical Conditions`, -n), y = n, fill = `Environmental Factors`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Most common medical conditions in people with hair loss",
    x = "Medical Condition",
    y = "Number of people",
    fill = "Environmental Factors"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))   # Rotate x-axis labels 45 degrees for readability

```

In general, the most common medical conditions among people with hair loss are Alopecia Areata, Androgenetic Alopecia, Psoriasis and Seborrheic Dermatitis. Environmental factors seem to have a slightly greater impact on conditions related to the skin or scalp, but are not the main factor in most cases.

### Boxplot relating age to hair loss — what can we conclude?

```{r}

# Create the boxplot with yellow fill and appropriate labels
boxplot(Age ~ `Hair_Loss_Label`, data = HairFallH_clean, col = "yellow",
        ylab = "Age", xlab = "Hair Loss",
        main = "Boxplot showing the relationship between age and hair loss") 

```

The boxplot shows that people without hair loss have a higher median age (around 35 years) and a larger box, indicating greater age dispersion in this group. In contrast, people with hair loss show a slightly lower median (around 34 years) and a narrower box, suggesting that ages are more concentrated around the median. Therefore, in this sample, people without hair loss tend to be slightly older and show more age variability than those who do experience hair loss.

### Frequency table and chart relating medical conditions to hair loss

```{r}

library(dplyr) 

# Create a frequency table for each combination of hair loss and medical condition
freq_table <- HairFallH_clean %>%
  group_by(Hair_Loss_Label, `Medical Conditions`) %>%  
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n))

freq_table

# Heat map
library(ggplot2)

ggplot(freq_table, aes(x = Hair_Loss_Label, y = `Medical Conditions`, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(
    title = "Heat map: Medical conditions according to hair loss",
    x = "Hair Loss",
    y = "Medical Conditions",
    fill = "Number of people"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Age histogram for people with hair loss, with normal density curve and normality assessment

```{r}

library(dplyr)
library(ggplot2)

# Filter only people with hair loss
hair_loss_data <- HairFallH_clean %>%
  filter(Hair_Loss_Label == "Hair Loss")
  
# Calculate mean and standard deviation of age
mean_age <- mean(hair_loss_data$Age)
sd_age   <- sd(hair_loss_data$Age)
  
# Create age range for the normal curve
age_range <- seq(min(hair_loss_data$Age),
                 max(hair_loss_data$Age),
                 length.out = 100)
  
# Create data frame for the normal curve
normal_curve <- data.frame(
  x = age_range,
  y = dnorm(age_range, mean = mean_age, sd = sd_age)
)

# Plot histogram + normal curve
ggplot(hair_loss_data, aes(x = Age)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightyellow", color = "black") +
  geom_density(color = "darkblue", linewidth = 1.2) +
  geom_line(data = normal_curve, aes(x = x, y = y),
            color = "red", linewidth = 1.2, linetype = "dashed") +
  labs(
    title = "Histogram of the age of people with hair loss",
    x = "Age",
    y = "Density"
  ) +
  theme_minimal()

# Shapiro-Wilk normality test
shapiro_result <- shapiro.test(hair_loss_data$Age)

# Q-Q plot to visualise normality
qqnorm(hair_loss_data$Age)
qqline(hair_loss_data$Age, col = "red")

# Numerical results
cat("Shapiro-Wilk test p-value:", round(shapiro_result$p.value, 6), "\n")

# Interpretation
if (shapiro_result$p.value < 0.05) {
  cat("Conclusion: The Q-Q plot may give the impression that normality is acceptable because the central part follows the line, but the formal test detects significant deviations (p < 0.05). This means we reject the null hypothesis of normality: the distribution differs significantly from a normal distribution.\n")
} else {
  cat("Conclusion: Normality cannot be rejected (p >= 0.05). Ages approximately follow a normal distribution.\n")
}
```

The histogram shows two density curves: the red curve represents the theoretical normal distribution calculated from the sample mean and standard deviation, while the blue curve corresponds to the empirical density estimated from the actual data. In this case, the blue curve does not fully coincide with the red one — it shows slight deviations and possible bimodality, suggesting that the age distribution of people with hair loss does not follow a perfectly normal shape.

The Q-Q plot shows that the points approximately follow the normality line in the central region, with visible deviations at the tails. A Shapiro-Wilk test is performed below to confirm.
